Scalable Extreme Deconvolution

### 2019-09-09 ml4ps_paper/paper.tex
Remove times from results table.

Initialisation of all models was performed by first running a minibatch version of the $K$-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
% TODO -- is data standardized?
Training was done for 20 epochs.
The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
We used a validation set comprising $10\%$ of the rows when developing our experiments.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method along with average training time.
Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$ and training time as function of mixture components $K$.
Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.

\begin{table}{}
  \caption{Average validation and test log-likelihoods with average training time for the Gaia data subset for different numbers of mixture components $K$.  Average over 10 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lcccc}
      \toprule
      Method     & K &  Validation     & Test & Time (minutes) \\
      \midrule
      Existing EM & 64 & $-26.10 \pm 0.03$ & - & $55.53 \pm 7.17$ \\
      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - & $100.30 \pm 1.96$ \\
       & 256 & $-25.76 \pm 0.02$ & - & $195.45 \pm 3.93$ \\
       & 512 & $-25.67 \pm 0.01$ & - & $385.60 \pm 8.24$ \\
      \midrule
      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - & $30.98 \pm 0.51$ \\
       & 128 & $-25.91 \pm 0.01$ & - & $47.01 \pm 0.67$ \\
       & 256 & $-25.83 \pm 0.00$ & - & $55.18 \pm 2.89$ \\
       & 512 & $-25.80 \pm 0.00$ & - & $100.47 \pm 3.89$ \\
      \midrule
      SGD & 64 & $-25.89 \pm 0.02$ & - & $28.02 \pm 0.77$ \\
       & 128 & $-25.77 \pm 0.02$ & - & $36.44 \pm 0.44$ \\
       & 256 & $-25.67 \pm 0.02$ & - & $40.54 \pm 1.48$ \\
       & 512 & $-25.59 \pm 0.02$ & - & $64.77 \pm 2.91$ \\
      \bottomrule
  \end{tabular}
\end{table}

>>>>

Initialisation of all models was performed by first running a minibatch version of the $K$-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
% TODO -- is data standardized?
Training was done for 20 epochs.
The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
We used a validation set comprising $10\%$ of the rows when developing our experiments.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$ and training time as function of mixture components $K$.
Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.

\begin{table}{}
  \caption{Average validation log-likelihoods for the Gaia data subset for different numbers of mixture components $K$, with average test log-likelihood for the best value of $K$ by validation. Average over 10 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lcccc}
      \toprule
      Method     & K &  Validation     & Test\\
      \midrule
      Existing EM & 64 & $-26.10 \pm 0.03$ & - \\
      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - \\
       & 256 & $-25.76 \pm 0.02$ & - \\
       & 512 & $-25.67 \pm 0.01$ & - \\
      \midrule
      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - \\
       & 128 & $-25.91 \pm 0.01$ & - \\
       & 256 & $-25.83 \pm 0.00$ & - \\
       & 512 & $-25.80 \pm 0.00$ & - \\
      \midrule
      SGD & 64 & $-25.89 \pm 0.02$ & - \\
       & 128 & $-25.77 \pm 0.02$ & - \\
       & 256 & $-25.67 \pm 0.02$ & - \\
       & 512 & $-25.59 \pm 0.02$ & - \\
      \bottomrule
  \end{tabular}
\end{table}
